Generalization Errors of Hard Parameter Sharing: High-dimensional Asymptotics and Implications

### 2020-10-03 arxiv/proof_examples.tex
revise

\subsubsection{Proof for Sample Imbalance}  
We now give the proof of Proposition \ref{cor_MTL_loss}. As in the proof of Theorem \ref{thm_many_tasks}, our proof is divided into the following steps: 
\begin{itemize}
\item simplifying the optimization objective using concentration estimates;

@@

In the specific setting in Proposition \ref{cor_MTL_loss}, we can further simplify $\cal L(\hat A)$ using Lemma \ref{lem_hat_v} and $\Sigma^{(1)}=\Sigma^{(2)}=\id_p$.

\begin{lemma}

@@

\end{lemma}

@@

\begin{proof}[Proof of Proposition \ref{cor_MTL_loss}]
%The proof of Lemma \ref{prop_model_shift_tight} is similar to that for Lemma \ref{thm_model_shift}, except that we can replace equation \eqref{bounddelta-} with a tighter bound. We only describe the main difference in the following proof.
%
%For any fixed $v\in \R$, $\beta_1 - v\beta_2$ is a random vector with i.i.d. entries with mean $0$ and variance $(1-v)^2\kappa^2 + d^2$. Then using the concentration result, Lemma \ref{largedeviation}, we get that for any constant $\e>0$, 
%\begin{align}
%&\left|\delta_{\bias}(v)- [(1-v)^2\kappa^2 + d^2]\tr (\cal K^\top \cal K)\right| \nonumber\\
%&=  \left|(\beta_1 - {v} \beta_2)^\top \cal K^\top \cal K (\beta_1 - {v} \beta_2) - [(1-v)^2\kappa^2 + d^2]\tr (\cal K^\top \cal K)\right| \nonumber\\
%&\le p^\e [(1-v)^2\kappa^2 + d^2] \left\{\tr\left[(\cal K^\top \cal K)^2\right]\right\}^{1/2} \lesssim p^{1/2+\e} [(1-v)^2\kappa^2 + d^2] \quad \text{w.h.p.},\label{anotherbeta}
%\end{align}
%where $\delta_{\bias}(v)$ was defined in equation \eqref{revise_deltabias}, $\cal K$ was defined as $\cal K:=v\Sigma_2^{1/2}({v}^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-1} X_1^{\top}X_1$, and in the last step we used $\|\cal K\|=\OO(1)$ by Fact \ref{lem_minv}. Now for $\tr (\cal K^\top \cal K)$, we rewrite it as
%\begin{align*}
%v^2 \tr \left[ ({v}^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}\Sigma_2({v}^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}(X_1^\top X_1)^2 \right].
%\end{align*}

Bounding $(X_1^\top X_1)^2=(Z_1^\top Z_1)^2$ using Fact equation \eqref{lem_minv}, we obtain that
\be\label{anotherwtbeta} \delta^\star_{\bias}(v) \cdot (\al_-^2(\rho_1)-\OO(p^{-1/2+\e})) \le  [(1-v)^2\kappa^2 + d^2]\tr (\cal K^\top \cal K) \le \delta^\star_{\bias}(v) \cdot  (\al_+^2(\rho_1)+\OO(p^{-1/2+\e})),\ee
where
$$\delta^\star_{\bias}(v):=n_1^2 v^2 [(1-v)^2\kappa^2 + d^2]\tr \left[ ({v}^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}\Sigma_2({v}^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}\right].$$
Note that $\delta^\star_{\bias}(v) \sim p$ by Fact \ref{lem_minv}. Hence combining equation \eqref{anotherbeta} and equation \eqref{anotherwtbeta} we get
\be\label{replaceest}
  \delta^\star_{\bias}(v) \cdot (\al_-^2(\rho_1)-\OO(p^{-1/2+\e})) \le  \delta_{\bias}(v)\le \delta^\star_{\bias}(v) \cdot  (\al_+^2(\rho_1)+\OO(p^{-1/2+\e})).
\ee
Now we can replace the estimate equation \eqref{bounddelta-} with this stronger estimate, and repeat all the other parts of the proof of Lemma \ref{thm_model_shift} to conclude Lemma \ref{prop_model_shift_tight}. In particular, one can calculate $  \delta^\star_{\bias}(v)$ using Lemma \ref{lem_cov_derivative} and get the $  \Delta^\star_{\bias}(v)$ term.

In the setting of Proposition \ref{prop_dist_transition}, we have $M = \Sigma_1^{1/2}\Sigma_2^{-1/2} = \id$. Then solving equations equation \eqref{eq_a12extra} and equation \eqref{eq_a34extra} with $\lambda_i\equiv 1$, we get that
	\begin{align}
		 a_1 = \frac{\rho_1(\rho_1 + \rho_2 - 1)}{(\rho_1 + \rho_2)^2} ,\quad
		& a_2 = \frac{\rho_2(\rho_1 + \rho_2 - 1)}{(\rho_1 + \rho_2)^2} , \label{simplesovlea12}\\
		a_3 = \frac{\rho_1}{(\rho_1 + \rho_2)(\rho_1 + \rho_2 - 1)}, \quad
		&  a_4 = \frac{\rho_2}{(\rho_1 + \rho_2)(\rho_1 + \rho_2 - 1)}.\label{simplesovlea34}
	\end{align}
	Using Lemma \ref{lem_minv} and Lemma \ref{lem_cov_shift}, we can track the reduction of variance from $\hat{\beta}_2^{\MTL}$ to $\hat{\beta}_2^{\STL}$ as 
\be\label{Deltavar}
\begin{split}
\delta_{\vari}&:=\sigma^2  \bigtr{(X_2^{\top}X_2)^{-1} }  - (1+\OO(n^{-\e}))\cdot \sigma^2  \bigtr{(X_1^{\top}X_1  + X_2^{\top}X_2)^{-1} } =\Delta_{\vari}\cdot (1+\OO(n^{-\e})) 
\end{split}
\ee
with high probability, where 
	\begin{align*}
		\Delta_{\vari} &\define \sigma^2 \bigbrace{\frac{1}{\rho_2 - 1} - \frac{1}{\rho_1 + \rho_2}\cdot\frac{1}{a_1+ a_2} } =\sigma^2  \cdot \frac{\rho_1}{(\rho_2-1)(\rho_1 + \rho_2 -1)}.
	\end{align*}
	%where we use equation \eqref{eq_a12} and Lemma \ref{lem_hat_v}.
	Next for the model shift bias
	$$\delta_{\bias}:=(1+\OO(n^{-\e}))\cdot d^2 \tr\left[(X_1^{\top}X_1 +X_2^{\top}X_2)^{-2} (X_1^{\top}X_1)^2\right], $$
	we can get from Lemma \ref{prop_model_shift_tight} that
\be\label{Deltabeta} 
\al_-^2(\rho_1) - \oo(1)  \le \frac{\delta_{\bias}}{ \Delta_{\bias}} \le \al_+^2(\rho_1) +  \oo(1) , \ee
	where 
	$$\Delta_{\bias}:=pd^2 \cdot \frac{\rho_1^2}{(\rho_1+\rho_2)^2}  \frac{1 + a_3 + a_4}{(a_1 + a_2)^2}= pd^2 \cdot \frac{\rho_1^2 (\rho_1+\rho_2)}{(\rho_1 + \rho_2 - 1)^3}.$$

	will also use equation \eqref{apply derivlocal}
\end{proof}

>>>>

\subsubsection{Proof for Sample Imbalance}  
We now give the proof of Proposition \ref{cor_MTL_loss}. Similar to the proof of Theorem \ref{thm_many_tasks}, our proof is divided into the following steps: 
\begin{itemize}
\item simplifying the optimization objective using concentration estimates;

@@

In the specific setting in Proposition \ref{cor_MTL_loss}, we can further simplify $\cal L(\hat A)$ using Claim \ref{lem_hat_v} and $\Sigma^{(1)}=\Sigma^{(2)}=\id_p$.

\begin{claim}

@@

\end{claim}

@@

\begin{proof}[Proof of Proposition \ref{cor_MTL_loss}]
%The proof of Lemma \ref{prop_model_shift_tight} is similar to that for Lemma \ref{thm_model_shift}, except that we can replace equation \eqref{bounddelta-} with a tighter bound. We only describe the main difference in the following proof.
%For any fixed $v\in \R$, $\beta_1 - v\beta_2$ is a random vector with i.i.d. entries with mean $0$ and variance $(1-v)^2\kappa^2 + d^2$. Then using the concentration result, Lemma \ref{largedeviation}, we get that for any constant $\e>0$, 
%\begin{align}
%&\left|\delta_{\bias}(v)- [(1-v)^2\kappa^2 + d^2]\tr (\cal K^\top \cal K)\right| \nonumber\\
%&=  \left|(\beta_1 - {v} \beta_2)^\top \cal K^\top \cal K (\beta_1 - {v} \beta_2) - [(1-v)^2\kappa^2 + d^2]\tr (\cal K^\top \cal K)\right| \nonumber\\
%&\le p^\e [(1-v)^2\kappa^2 + d^2] \left\{\tr\left[(\cal K^\top \cal K)^2\right]\right\}^{1/2} \lesssim p^{1/2+\e} [(1-v)^2\kappa^2 + d^2] \quad \text{w.h.p.},\label{anotherbeta}
%\end{align}
%where $\delta_{\bias}(v)$ was defined in equation \eqref{revise_deltabias}, $\cal K$ was defined as $\cal K:=v\Sigma_2^{1/2}({v}^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-1} X_1^{\top}X_1$, and in the last step we used $\|\cal K\|=\OO(1)$ by Fact \ref{lem_minv}. Now for $\tr (\cal K^\top \cal K)$, we rewrite it as
%\begin{align*}
%v^2 \tr \left[ ({v}^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}\Sigma_2({v}^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}(X_1^\top X_1)^2 \right].
%\end{align*}
%Bounding $(X_1^\top X_1)^2=(Z_1^\top Z_1)^2$ 
Using \eqref{lem_cov_shift_eq}, we obtain that 
\be\label{eq_var111}\bigtr{\Sigma(1)^{-1}  } = \tr\left[ \left((X^{(1)})^\top X^{(1)}  + (X^{(2)})^\top X^{(2)}\right)^{-1}\right]=\bigtr{\frac{(a_1\id_p+a_2\id_p)^{-1}}{n_{1}+n_2} }+\OO(p^{-c_\varphi})\ee
with high probability. Then solving equation \eqref{eq_a12extra} with $\lambda_i\equiv 1$, $1\le i\le p$, we get that with high probability,
	\begin{align}
		 a_1 = \frac{\rho_1(\rho_1 + \rho_2 - 1)}{(\rho_1 + \rho_2)^2} ,\quad
		& a_2 = \frac{\rho_2(\rho_1 + \rho_2 - 1)}{(\rho_1 + \rho_2)^2} . \label{simplesovlea12}
			\end{align}
Inserting it into \eqref{eq_var111}, we obtain that 
\be\label{eq_var112}\bigtr{\Sigma(1)^{-1}  } = \frac{p}{n_1+n_2} \cdot \frac{\rho_1+\rho_2}{\rho_1+\rho_2-1}+\OO(p^{-c_\varphi})=  \frac{1}{\rho_1+\rho_2-1}+\OO(p^{-c_\varphi})\ee
with high probability.

On the other hand, using Fact \ref{lem_minv} (iii), we can bound that with high probability,
\begin{align}
 \frac{(\sqrt{n_1}-\sqrt{p})^4 \cdot (1- p^{-c_\varphi})}{p}\sum_{i=1}^p \left(\Sigma(1)^{-2}\right)_{ii}  &\le {p}^{-1}\tr\left[\Sigma(1)^{-2} \left((X^{(1)})^\top X^{(1)} \right)^2\right] \label{eq_bias111}\\ 
 &\le \frac{(\sqrt{n_1}+\sqrt{p})^4 \cdot (1+ p^{-c_\varphi})}{p}\sum_{i=1}^p \left(\Sigma(1)^{-2}\right)_{ii} .\nonumber
\end{align}
To obtain this inequality, we used  
$$ (\sqrt{n_1}-\sqrt{p})^4 \cdot (1-p^{-c_\varphi}) \prec \left((X^{(1)})^\top X^{(1)} \right)^2 \preceq (\sqrt{n_1}+\sqrt{p})^4 \cdot (1+ p^{-c_\varphi}) \quad \text{with high probability},$$
by Fact \ref{lem_minv} (iii), and the fact that for the product of two PSD matrices, its trace is always nonnegative.
Using \eqref{lem_cov_derv_eq} with $\Sigma^{(1)}=\Sigma^{(2)}=\Lambda=V=\id_p$ and $w$ being the $i$-th coordinate vector, we can calculate that 
\begin{align}\label{eq_bias112}
 \left(\Sigma(1)^{-2}\right)_{ii} = \frac{1}{(n_1+n_2)^2}\cdot \frac{a_3+a_4+1}{(a_1+a_2)^2} +\OO(p^{-c_\varphi})
\end{align}
with high probability. Solving equation \eqref{eq_a34extra} with $a_1, a_2$ in \eqref{simplesovlea12} and $\lambda_i\equiv 1$, $1\le i\le p$, we can obtain that
	\begin{align}
		% a_1 = \frac{\rho_1(\rho_1 + \rho_2 - 1)}{(\rho_1 + \rho_2)^2} ,\quad & a_2 = \frac{\rho_2(\rho_1 + \rho_2 - 1)}{(\rho_1 + \rho_2)^2} , \label{simplesovlea12}\\
		a_3 = \frac{\rho_1}{(\rho_1 + \rho_2)(\rho_1 + \rho_2 - 1)}, \quad
		&  a_4 = \frac{\rho_2}{(\rho_1 + \rho_2)(\rho_1 + \rho_2 - 1)}.\label{simplesovlea34}
	\end{align}
Inserting them into \eqref{eq_bias112}, we obtain that 
\begin{align*}%\label{eq_bias113}
 \left(\Sigma(1)^{-2}\right)_{ii} = \frac{1}{(n_1+n_2)^2}\cdot \frac{(\rho_1+\rho_2)^3}{(\rho_1+\rho_2-1)^3} +\OO(p^{-c_\varphi})
\end{align*}
with high probability. In fact, we need this estimate to hold simultaneously for all $1\le i \le p$ with high probability. For this purpose, we shall use \eqref{apply derivlocal} to get that 
\begin{align*}%\label{eq_bias113}
\left| \left(\Sigma(1)^{-2}\right)_{ii} - \frac{1}{(n_1+n_2)^2}\cdot \frac{(\rho_1+\rho_2)^3}{(\rho_1+\rho_2-1)^3}\right|\prec p^{-\frac{\varphi-4}{2\varphi}}
\end{align*}
on a high probability event that does not depend on $i$. Then using Fact \ref{lem_stodomin} (i), we obtain that 
\begin{align*} 
\left| \sum_{i=1}^p \left(\Sigma(1)^{-2}\right)_{ii} - \frac{p}{(n_1+n_2)^2}\cdot \frac{(\rho_1+\rho_2)^3}{(\rho_1+\rho_2-1)^3}\right|\prec p^{1-\frac{\varphi-4}{2\varphi}}
\end{align*}
with high probability, which by Definition \ref{stoch_domination} gives that
\begin{align}\label{eq_bias113}
\left| p^{-1}\sum_{i=1}^p \left(\Sigma(1)^{-2}\right)_{ii} - \frac{1}{(n_1+n_2)^2}\cdot \frac{(\rho_1+\rho_2)^3}{(\rho_1+\rho_2-1)^3}\right|\le p^{-c_\varphi}
\end{align}
with high probability. Inserting \eqref{eq_bias113} into \eqref{eq_bias111}. we get
\begin{align}
& \frac{(\sqrt{n_1}-\sqrt{p})^4 \cdot (1- p^{-c_\varphi}) - n_1^2 \cdot (1+\OO(p^{-c_\varphi}))}{(n_1+n_2)^2}\cdot  \frac{(\rho_1+\rho_2)^3}{(\rho_1+\rho_2-1)^3}\nonumber \\
&\le  {p}^{-1}\tr\left[\Sigma(1)^{-2} \left((X^{(1)})^\top X^{(1)} \right)^2\right] - \frac{n_1^2}{(n_1+n_2)^2}\cdot  \frac{(\rho_1+\rho_2)^3}{(\rho_1+\rho_2-1)^3} \nonumber\\
&  \le \frac{(\sqrt{n_1}+\sqrt{p})^4 \cdot (1+ p^{-c_\varphi}) - n_1^2 \cdot (1+\OO(p^{-c_\varphi}))}{(n_1+n_2)^2}\cdot  \frac{(\rho_1+\rho_2)^3}{(\rho_1+\rho_2-1)^3}\label{eq_bias114}
\end{align}
with high probability.

Now plugging \eqref{eq_var112} and \eqref{eq_bias114} into \eqref{simple1}, we get
\begin{align}
\left|\cal L(\hat A)- -\frac{\sigma^2}{\rho_1+\rho_2-1} - d^2\cdot \frac{\rho_1^2 (\rho_1+\rho_2)}{(\rho_1 + \rho_2 - 1)^3}\right| \le   \left[\left( 1+\frac{1}{\sqrt{\rho_1}}\right)^4-1\right] d^2\cdot \frac{\rho_1^2 (\rho_1+\rho_2)}{(\rho_1 + \rho_2 - 1)^3} \nonumber\\
+C\left( p^{-c_\varphi} (\sigma^2 + d^2)+ \frac{d^4 + \sigma^2 d^2}{\kappa^2}+p^{-1/2+2e}\kappa^2 +  p^{-1/4+e} (\sigma^2 +d^2) \right) \label{simple1add}
\end{align}
with high probability for some constant $C>0$. Then using Claim \ref{claim_largedev2} and noticing that the constant $e>0$ can be arbitrarily small, we conclude \eqref{cor_MTL_error}.
\end{proof}
%\begin{align}
%				a_3 + a_4 = \frac{1}{n_1 + n_2}\sum_{i=1}^p \frac{1}{\lambda_i^2 a_1 + a_2},
%				a_3 + \frac{1}{n_1 + n_2} \sum_{i=1}^p \frac{\lambda_i^2 (a_1 a_4 - a_2 a_3)}{(\lambda_i^2 a_1 + a_2)^2} = \frac{1}{n_1 + n_2} \sum_{i=1}^p \frac{\lambda_i^2 a_1}{(\lambda_i^2 a_1 + a_2)^{2}}.
%%				\left(\frac{\rho_1}{a_1^{2}} -  b_2  \right)\cdot  a_3 -  b_1 \cdot  a_4 = b_1,\quad \left(\frac{\rho_2}{a_2^{2}}-  b_0\right)\cdot  a_4 - b_1 \cdot  a_3
%%				= b_0.
%			\end{align}
%\be\label{anotherwtbeta} \delta^\star_{\bias}(v) \cdot (\al_-^2(\rho_1)-\OO(p^{-1/2+\e})) \le  [(1-v)^2\kappa^2 + d^2]\tr (\cal K^\top \cal K) \le \delta^\star_{\bias}(v) \cdot  (\al_+^2(\rho_1)+\OO(p^{-1/2+\e})),\ee
%where
%$$\delta^\star_{\bias}(v):=n_1^2 v^2 [(1-v)^2\kappa^2 + d^2]\tr \left[ ({v}^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}\Sigma_2({v}^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}\right].$$
%Note that $\delta^\star_{\bias}(v) \sim p$ by Fact \ref{lem_minv}. Hence combining equation \eqref{anotherbeta} and equation \eqref{anotherwtbeta} we get
%\be\label{replaceest}
%  \delta^\star_{\bias}(v) \cdot (\al_-^2(\rho_1)-\OO(p^{-1/2+\e})) \le  \delta_{\bias}(v)\le \delta^\star_{\bias}(v) \cdot  (\al_+^2(\rho_1)+\OO(p^{-1/2+\e})).
%\ee
%Now we can replace the estimate equation \eqref{bounddelta-} with this stronger estimate, and repeat all the other parts of the proof of Lemma \ref{thm_model_shift} to conclude Lemma \ref{prop_model_shift_tight}. In particular, one can calculate $  \delta^\star_{\bias}(v)$ using Lemma \ref{lem_cov_derivative} and get the $  \Delta^\star_{\bias}(v)$ term.

%In the setting of Proposition \ref{prop_dist_transition}, we have $M = \Sigma_1^{1/2}\Sigma_2^{-1/2} = \id$. Then solving equations equation \eqref{eq_a12extra} and equation \eqref{eq_a34extra} with $\lambda_i\equiv 1$, we get that
%	\begin{align}
%		 a_1 = \frac{\rho_1(\rho_1 + \rho_2 - 1)}{(\rho_1 + \rho_2)^2} ,\quad
%		& a_2 = \frac{\rho_2(\rho_1 + \rho_2 - 1)}{(\rho_1 + \rho_2)^2} , \label{simplesovlea12}\\
%		a_3 = \frac{\rho_1}{(\rho_1 + \rho_2)(\rho_1 + \rho_2 - 1)}, \quad
%		&  a_4 = \frac{\rho_2}{(\rho_1 + \rho_2)(\rho_1 + \rho_2 - 1)}.\label{simplesovlea34}
%	\end{align}
%	Using Lemma \ref{lem_minv} and Lemma \ref{lem_cov_shift}, we can track the reduction of variance from $\hat{\beta}_2^{\MTL}$ to $\hat{\beta}_2^{\STL}$ as 
%\be\label{Deltavar}
%\begin{split}
%\delta_{\vari}&:=\sigma^2  \bigtr{(X_2^{\top}X_2)^{-1} }  - (1+\OO(n^{-\e}))\cdot \sigma^2  \bigtr{(X_1^{\top}X_1  + X_2^{\top}X_2)^{-1} } =\Delta_{\vari}\cdot (1+\OO(n^{-\e})) 
%\end{split}
%\ee
%with high probability, where 
%	\begin{align*}
%		\Delta_{\vari} &\define \sigma^2 \bigbrace{\frac{1}{\rho_2 - 1} - \frac{1}{\rho_1 + \rho_2}\cdot\frac{1}{a_1+ a_2} } =\sigma^2  \cdot \frac{\rho_1}{(\rho_2-1)(\rho_1 + \rho_2 -1)}.
%	\end{align*}
%	%where we use equation \eqref{eq_a12} and Lemma \ref{lem_hat_v}.
%	Next for the model shift bias
%	$$\delta_{\bias}:=(1+\OO(n^{-\e}))\cdot d^2 \tr\left[(X_1^{\top}X_1 +X_2^{\top}X_2)^{-2} (X_1^{\top}X_1)^2\right], $$
%	we can get from Lemma \ref{prop_model_shift_tight} that
%\be\label{Deltabeta} 
%\al_-^2(\rho_1) - \oo(1)  \le \frac{\delta_{\bias}}{ \Delta_{\bias}} \le \al_+^2(\rho_1) +  \oo(1) , \ee
%	where 
%	$$\Delta_{\bias}:=pd^2 \cdot \frac{\rho_1^2}{(\rho_1+\rho_2)^2}  \frac{1 + a_3 + a_4}{(a_1 + a_2)^2}= pd^2 \cdot \frac{\rho_1^2 (\rho_1+\rho_2)}{(\rho_1 + \rho_2 - 1)^3}.$$
